# Missing values

## Number of missing attributes

As we had a lot of data initially ~37M rows, we removed rows from each categorical variable, that were less in percentage. For more details, see the data transformation section [here](url). The following visualization gives information about missing data after the transformation:

```{r missing_value_load_imports, echo = FALSE}
library("rjson")
library("dplyr")
library(tidyverse)
library(ggplot2)
library(dplyr)
library(tidyverse)
library(choroplethr)
library(statebins)
library(maps)
library(ggrepel)
library(reshape2)
library(viridis)
library(ggthemes)
library(gapminder)
library(ggthemes)
library(gganimate)
library(ggmosaic)
library(scales)
load("data/tedsa_puf_2000_2019.RData")

```


```{r missing_value_analysis,fig.height = 6,fig.width = 10, echo = FALSE}
missing_value_patterns <- function(df, percent = TRUE) {
  df[df=="Missing"] <- NA
  if (percent == TRUE) {
    var = 100 / dim(df)[1]
    y_label = "% rows missing"
    x_label = "Percentage of rows"
  }
  else
  {
    var = 1
    y_label = "rows missing"
    x_label = "rows count"
  }
  
  missing_patterns <- data.frame(is.na(df)) %>%
    group_by_all() %>%
    count(name = "count", sort = TRUE) %>%
    ungroup()
  
  missing_patterns <-
    missing_patterns %>% mutate(missing_pattern = seq(1, dim(missing_patterns)[1]))
  
  missing_patterns$complete_case <-
    missing_patterns[, -c(dim(missing_patterns)[2] - 1, dim(missing_patterns)[2]), drop = FALSE] %>% rowSums()
  
  missing_patterns$complete_case <-
    !as.logical(missing_patterns$complete_case)
  
  missing_patterns_pivot <-
    missing_patterns %>% select(-c(count)) %>% pivot_longer(!c(missing_pattern, complete_case),
                                                            names_to = 'variable',
                                                            values_to = 'value')
  
  column_wise = data.frame(na_count = colSums(is.na(df)))
  column_wise$variable = row.names(column_wise)
  
  order_of_variables = column_wise %>% arrange(desc(na_count)) %>% pull(variable)
  
  p1 <-
    ggplot(column_wise, aes(
      x = factor(variable, levels = order_of_variables),
      y = na_count * var
    )) + geom_bar(stat = 'identity',
                  fill = '#98B7F3',
                  alpha = 0.6) + xlab('') + ylab(y_label) + theme_bw() + theme(panel.grid.major.x = element_blank())  +   theme(axis.text.x = element_text(angle = 60),
                                                                                                                                plot.title = element_text(hjust = 0.5))
  
  if (percent == TRUE) {
    p1 <-
      p1 + scale_y_continuous(
        expand = c(0, 0),
        breaks = pretty_breaks(),
        limits = c(0, 100)
      )
  }
  else
  {
    p1 <-
      p1 + scale_y_continuous(expand = c(0, 0), breaks = pretty_breaks())
  }
  
  
  p2 <-
    ggplot(missing_patterns_pivot,
           aes(
             factor(variable, levels = order_of_variables),
             factor(reorder(
               missing_pattern, desc(missing_pattern)
             )),
             fill = value,
             alpha = complete_case
           )) + geom_tile(color = "white") + scale_y_discrete(expand = c(0, 0)) + scale_x_discrete(expand = c(0, 0)) + theme_bw() + theme (
             legend.position = "None",
             panel.grid.major.x = element_blank(),
             panel.grid.major.y = element_blank()
           ) + scale_fill_manual(values = c("#CBCBCB", "#B69EE6")) + ylab('missing pattern') + xlab('variable') + scale_alpha_manual(values = c(0.6, 1)) + geom_text(
             x = floor(dim(column_wise)[1] / 2),
             y = dim(missing_patterns)[1] + 1 - missing_patterns[missing_patterns$complete_case == TRUE, "missing_pattern"] %>% pull(),
             label = "Complete cases",
             size = 3
           )  +   theme(axis.text.x = element_text(angle = 60),
                        plot.title = element_text(hjust = 0.5))
  
  p3 <-
    ggplot(missing_patterns,
           aes(
             y = factor(reorder(
               missing_pattern, desc(missing_pattern)
             )),
             x = count * var,
             alpha = complete_case
           )) + geom_bar(stat = 'identity', fill = '#98B7F3') + ylab('') + xlab(x_label) +
    theme_bw() + theme(panel.grid.major.y = element_blank(),
                       legend.position = "None") + scale_alpha_manual(values = c(0.6, 1)) + scale_y_discrete(expand = c(0, 0))  +   theme(axis.text.x = element_text(angle = 60),
                                                                                                                                          plot.title = element_text(hjust = 0.5))
  
  if (percent == TRUE) {
    p3 <-
      p3 + scale_x_continuous(
        expand = c(0, 0),
        breaks = pretty_breaks(),
        limits = c(0, 100)
      )
  }
  else
  {
    p3 <-
      p3 + scale_x_continuous(expand = c(0, 0), breaks = pretty_breaks())
  }
  
  
  p1 + plot_spacer() + p2 + p3 + plot_layout(
    nrow = 2,
    ncol = 2,
    widths = c(8, 2),
    heights = c(2, 8)
  ) + plot_annotation(title = 'Missing value patterns')
}

missing_value_patterns(dataset)
```
Observations:

1. Only the Employment column seem to be missing values. About 10% of rows in the data seem to have missing value for this column. We feel 10% is a lot of data to be removed directly and it might still hold information. For example: The information is missing for a particular state or year etc.

## Understanding missing values pattern

```{r missing_pattern_analysis,fig.height = 6,fig.width = 10, echo = FALSE}
missing <- dataset %>% filter(EMPLOY == "Missing")
missing_state <- missing %>% group_by(STFIPS) %>% summarise(`count` = n())
missing_state['Percentage_of_rows'] <- (missing_state['count']/dim(missing)[1]) * 100 
g1 <- ggplot(missing_state[1:20, ], aes(x = Percentage_of_rows, y = reorder(STFIPS, Percentage_of_rows))) + geom_point() + labs(x = 'Percentage of Missing Value', y = "State", title = "Percentage of missing employment data by states") + theme_bw()

missing_year <- missing %>% group_by(ADMYR) %>% summarise(`count` = n())
missing_year['Percentage_of_rows'] <- (missing_year['count']/dim(missing)[1]) * 100 
g2 <- ggplot(missing_year[1:20, ], aes(x = Percentage_of_rows, y = ADMYR)) + geom_point() + labs(x = 'Percentage of Missing Value', y = "Year", title = "Percentage of missing employment data by years") + theme_bw()
g1+g2
```
Onservations:

1. The Employment data was missing during the early 2000's. The recent data has information about employment.
2. The state of California account for 30% of the missing data. This is much larger than any other state.



